Abstraction and Invariance for Algebraically Indexed Types

### 2012-07-09 paper/section-introduction.tex
Patty's changes

\section{Introduction}
\label{sec:introduction}
The best way we know of describing the semantics of parametric
polymorphism is \emph{relational parametricity}, whose central result
is Reynolds' Abstraction Theorem~\cite{reynolds83types}. Its striking
consequences include the well-known ``theorems for free'' for
polymorphic types~\cite{wadler89theorems}, non-inhabitation results, and precise
correspondences between System F encodings and algebraic datatypes~\cite{PittsAM:parpoe},
abstract data types, and, most recently, higher-order encodings of
binder syntax~\cite{syntaxforfree}.

Relational parametricity is in essence a principle of
\emph{invariance}: the behaviour of polymorphic code is invariant
under changes to data representation. Invariance results also abound in
mathematics and physics. The area of a triangle is invariant with
respect to isometries of the Euclidean plane; the determinant of a
matrix is invariant under changes of basis; and Newton's laws are the
same in all inertial frames. Typically, transformations in mathematics
and physics have interesting structure; for example, translations in
the Euclidean plane form an Abelian group.

%Types express properties of programs. The statement $\mathit{b}: \tyBool$ tells
%us that $\mathit{b}$ has value $\mathtt{true}$ or $\mathtt{false}$;
%the statement $\mathit{pos}:\tyReal\tyProduct\tyReal$ says that
%$\mathit{pos}$ is a pair whose components are real numbers. More interestingly,
%the statement $\mathit{f} : \forall X. \tyPrim{list}X\tyArr\tyPrim{list}X$ tells us that
%if we apply $\mathit{f}$ to a list whose elements satisfy some property $\psi$, 
%then the result is a list whose elements also satisfy $\psi$.

%Unary predicates only go so far, and for polymorphic types
%particularly do not express everything that follows from possessing that type. 
%Imagine a function $\mathit{oddRev}$ that reverses any list whose first element
%is an odd number, otherwise behaving as the identity. It certainly
%preserves properties of the elements -- but truly parametric
%polymorphism should tell us more.

%To obtain stronger properties we must \emph{relate} multiple
%executions of the same code. The statement $\mathit{f}:\forall
%X. \tyPrim{list}X\tyArr\tyPrim{list}X$ then says that any change in
%representation of the list elements supplied to $\mathit{f}$ is
%reflected in the elements in the result.  Formally, for any function
%$g$, we have $\mathit{f}(\mathrm{map}\:\mathit{g}\:\mathit{x}) =
%\mathrm{map}\:\mathit{g}\:(\mathit{f}\:\mathit{x})$.  This is the
%essence of \emph{relational parametricity} as formulated by
%Reynolds~\cite{reynolds83types}. Its consequences are striking.  It
%immediately tells us that $\mathit{oddRev}$ cannot be given a
%polymorphic type: simply consider the change-of-representation given
%by $\lambda x. x+1$.  But more profoundly it can be used to prove blah
%blah (initial algebras, term reps, non-inhabitation, etc.).

Inspired by this connection, we study relational parametricity for
types that are \emph{indexed} by attributes that themselves have
algebraic structure.  For example, in computational geometry, points
in 2-space can be indexed by a coordinate frame; in physical
modelling, quantities are indexed by units-of-measure; in
information-flow security, computations are indexed by principals.
Types that are polymorphic over such indices induce invariance
properties, and introduce a useful abstraction barrier.

\paragraph{Invariance}
A function $\mathit{areaTri}$ that computes the area of a triangle might
have simple type $\tyPrimNm{vec}\times\tyPrimNm{vec}\times\tyPrimNm{vec}\tyArr\tyReal$
but we can assign it a more expresive polymorphic type
\[
\mathrm{areaTri} : \forall\alpha\mathord:\SynTransl{2}.
  \tyPrim{vec}{\alpha} \times \tyPrim{vec}{\alpha} \times \tyPrim{vec}{\alpha} \to \tyReal \\
\]
in which the index $\alpha$ can be thought of as a `translation frame' that is
preserved by the vector arguments, from which we can deduce that
\[
\mathrm{areaTri}\;(t + v_1, t + v_2, t + v_3) = 
\mathrm{areaTri}\;(v_1, v_2, v_3)
\]
for any translation $t\in\Transl{2}$.

In more complex examples the invariants \emph{compose} in various
ways; moreover, the compositions satisfy algebraic laws. Consider a function
that computes the area of a circle given its radius; by ascribing it
the polymorphic type
\[
\mathrm{areaCircle} : \forall\alpha\mathord:\SynScal.\tyPrim{real}{\alpha}\to
\tyPrim{real}{\alpha\cdot\alpha}
\]
we have captured that its area varies as the square of its radius, that is, for any $s>0$,
\[
\mathrm{areaCircle}\;(s\cdot r) = s^2\cdot \mathrm{areaCircle}\;r.
\]
Here $\alpha$ can be interpreted as the \emph{units of measure} of the argument
to $\mathrm{areaCircle}$, and `$\cdot$' composes units by forming
their product. To complete the system, we would add an inverse
operation and identity unit-of-measure, together with the algebraic laws of
Abelian groups, permitting, for example, the identification of
$\tyPrim{real}{\alpha\cdot\alpha^{-1}}$ with $\tyPrim{real}{1}$.

\paragraph{Abstraction}
In his original paper on parametricity, Reynolds asserted that \emph{type structure is
  a syntactic discipline for enforcing levels of abstraction}.  We see
something analogous here: supposing that all primitive operations are
given types that reflect their behaviour under translation, there is
no way to `break' this property, for example, for $\mathrm{areaTri}$
to depend on the actual coordinates of its inputs. Furthermore, the
distinction between points and vectors that is often enforced through
abstract data types~\cite{CGAL} is here captured by the indices
instead. For example, the operation that takes two points and computes
their vector `difference' can be assigned the type
$\forall\alpha\mathord:\SynTransl.\tyPrim{vec}\alpha\times\tyPrim{vec}\alpha\to\tyPrim{vec}0$,
reflecting the invariance of the result (a pure vector) under
translations of the point arguments. In essence, through types alone
we can derive so-called \emph{coordinate-free} geometry~\cite{CFGADT}.

The invariance properties listed above can be seen as `theorems for
free'~\cite{wadler89theorems} obtained from types alone. We can
further observe that the abstraction afforded by polymorphic indexed
types induces interesting type \emph{isomorphisms}.  The type of
$\mathrm{areaCircle}$ above is in fact isomorphic to
$\tyPrim{real}{1}$, that is, a dimensionless constant. A moment's
thought reveals why: what possible unary functions can be constructed whose
outputs scale as the square of the scaling of their inputs? Answer:
just those functions of the form~$\lambda x. k x^2$ for some constant~$k$.
In this case, of course, we expect that~$k = \pi$.

\paragraph{Relational parametricity}
In order to derive such invariance and abstraction properties of types, we
adopt the techniques of relational parametricity. Over an underlying index-erasure semantics
we construct binary
relations parameterized by an environment $\rho$ that describes how values of primitive type
are related according to their indices. 
For example, values $v$ and
$w$ of type $\tyPrim{real}\alpha$ are related when $v$ `scales to' $w$
according to an interpretation of $\alpha$ (that is, $w=\rho(\alpha)\cdot v$).
Values of polymorphic type are related exactly when they are related for
all possible interpretations of the quantified variable. For example, 
values $v$ and $w$ of type $\forall\alpha\mathord:\SynTransl.\tyPrim{vec}\alpha\to
\tyPrim{vec}\alpha$ are related when they are related at type 
$\tyPrim{vec}\alpha\to\tyPrim{vec}\alpha$ for all 
translations $t\in\Transl{2}$ associated with $\alpha$.

As it happens, these examples can be expressed using unary functions
rather than binary relations: we can write functions that scale, or
translate, one value to another according to their type, even for
higher-order function types. Also note that the interpretation of
indices can be built up compositionally, by recursion over the
syntactic structure. For example, we have $\rho(\alpha\cdot\alpha) =
\rho(\alpha)\cdot\rho(\alpha)$ where product of units-of-measure is interpreted by
product of scale factors.

For more sophisticated applications we need the fully generality of
binary relations. First, as is also the case with type polymorphism,
it provides the appropriate meaning for nested quantifiers in types
such as
$(\forall\alpha\mathord:\SynScal.\tyPrim{real}\alpha\to\tyPrim{real}\alpha)\to\tyBool$.
More interestingly, we sometimes use relations that are not built up
in a compositional way -- we can, for instance, use such relations to
prove that there are no non-trivial functions of type
$\forall\alpha\mathord:\SynScal.\tyPrim{real}{\alpha\cdot\alpha}\to\tyPrim{real}\alpha$.
Finally, there are interesting applications in which even the basic
relations are not functional. Consider a type system in which the
index in $\tyPrim{real}\alpha$ is interpreted not as a
unit-of-measure, but as a measure of \emph{closeness}. Two values $x$
and $y$ of this type are related if $|x-y| < \rho(\alpha)$, in which
$\rho(\alpha)$ is a positive real number.  Rather beautifully, the
standard notion of continuous functions can then be expressed as a
type:
\begin{displaymath}
  \forall \epsilon \mathord: \mathsf{R}^{>0}.\ \exists \delta\mathord: \mathsf{R}^{>0}.\ \tyPrim{real}{\delta} \to \tyPrim{real}{\epsilon}
\end{displaymath}

@@

We make the following contributions. 
\begin{itemize}
\item 
We present a collection of compelling examples of algebraically indexed
types, including a novel type system for geometry, a type system for
tracking taint marking (or provenance) of strings, a refined type
system for information flow, based on logic, and a simple type
system for basic continuity analysis.
\item 
We formulate a general class of type systems that can either be
used as programming languages in their own right, or as the targets of
type-based analyses. The type system consists of the usual type
constructors together with a collection of indexed primitive types,
universal and existential quantification over the indices, and a
multi-sorted equational theory for indices. 
\item
We describe a relational semantics for the type system and prove an analogue
of Reynolds' Abstraction Theorem. We pin down precise conditions on sets of
relation environments for which the theorem holds.
\item
For each of our main examples we deduce free theorems that are consequences of
the Abstraction Theorem, prove non-definability results, and derive interesting
type isomorphisms.
\end{itemize}
We have fully formalized the framework and Abstraction Theorem using the Coq proof assistant,
making use of strongly-typed term representations throughout~\cite{TypedSyntax}.

>>>>

\section{Introduction}
\label{sec:introduction}
The best way we know of describing the semantics of parametric
polymorphism is \emph{relational parametricity}, whose central result
is Reynolds' Abstraction Theorem~\cite{reynolds83types}. Its striking
consequences include the well-known ``free theorems'' for polymorphic
types~\cite{wadler89theorems}, non-inhabitation results, and precise
correspondences between System F encodings and algebraic
datatypes~\cite{PittsAM:parpoe}, abstract data types, and, most
recently, higher-order encodings of binder
syntax~\cite{syntaxforfree}.

Relational parametricity is in essence a principle of
\emph{invariance}: the behaviour of polymorphic code is invariant
under changes of data representation. Invariance results also abound in
mathematics and physics. The area of a triangle is invariant with
respect to isometries of the Euclidean plane; the determinant of a
matrix is invariant under changes of basis; and Newton's laws are the
same in all inertial frames. Typically, transformations in mathematics
and physics have interesting structure; for example, translations in
the Euclidean plane form an abelian group.

%Types express properties of programs. The statement $\mathit{b}:
%\tyBool$ tells us that $\mathit{b}$ has value $\mathtt{true}$ or
%$\mathtt{false}$; the statement
%$\mathit{pos}:\tyReal\tyProduct\tyReal$ says that $\mathit{pos}$ is a
%pair whose components are real numbers. More interestingly, the
%statement $\mathit{f} : \forall
%X. \tyPrim{list}X\tyArr\tyPrim{list}X$ tells us that if we apply
%$\mathit{f}$ to a list whose elements satisfy some property $\psi$,
%then the result is a list whose elements also satisfy $\psi$.

%Unary predicates only go so far, and for polymorphic types
%particularly do not express everything that follows from possessing
%that type.  Imagine a function $\mathit{oddRev}$ that reverses any
%list whose first element is an odd number, otherwise behaving as the
%identity. It certainly preserves properties of the elements -- but
%truly parametric polymorphism should tell us more.

%To obtain stronger properties we must \emph{relate} multiple
%executions of the same code. The statement $\mathit{f}:\forall
%X. \tyPrim{list}X\tyArr\tyPrim{list}X$ then says that any change in
%representation of the list elements supplied to $\mathit{f}$ is
%reflected in the elements in the result.  Formally, for any function
%$g$, we have $\mathit{f}(\mathrm{map}\:\mathit{g}\:\mathit{x}) =
%\mathrm{map}\:\mathit{g}\:(\mathit{f}\:\mathit{x})$.  This is the
%essence of \emph{relational parametricity} as formulated by
%Reynolds~\cite{reynolds83types}. Its consequences are striking.  It
%immediately tells us that $\mathit{oddRev}$ cannot be given a
%polymorphic type: simply consider the change-of-representation given
%by $\lambda x. x+1$.  But more profoundly it can be used to prove
%blah blah (initial algebras, term reps, non-inhabitation, etc.).

In this paper we show how to capture, and generalise, invariance under
groups of translations using
%Inspired by this connection, we study 
relational parametricity for types that are indexed by attributes with
algebraic structure. For example, in computational geometry, points in
2-space can be indexed by a coordinate frame; in physical modelling,
quantities are indexed by units-of-measure; in information-flow
security, computations are indexed by principals.  Types that are
polymorphic over such indices induce invariance properties and
abstraction barriers beyond those introduced by their unindexed
versions, as we now illustrate.

\paragraph{Invariance}
A function $\mathit{areaTri}$ that computes the area of a triangle
can be assigned the type:
$\tyPrimNm{vec}\times\tyPrimNm{vec}\times\tyPrimNm{vec}\tyArr\tyReal$.
But we can also assign it the following more expressive polymorphic
type:
\[
\mathrm{areaTri} : \forall\alpha\mathord:\SynTransl{2}.
  \tyPrim{vec}{\alpha} \times \tyPrim{vec}{\alpha} \times \tyPrim{vec}{\alpha} \to \tyReal \\
\]
in which the index $\alpha$ is thought of as a ``translation frame''
that is preserved by the vector arguments. From this we can deduce
that
\[
\mathrm{areaTri}\;(t + v_1, t + v_2, t + v_3) = 
\mathrm{areaTri}\;(v_1, v_2, v_3)
\]
for any translation $t\in\Transl{2}$.

In more complex examples the invariants \emph{compose} in various
ways, with the compositions satisfy algebraic laws. For example, 
we can assign a function that computes the area of a circle given its
radius the following polymorphic type:
\[
\mathrm{areaCircle} : \forall\alpha\mathord:\SynScal.\tyPrim{real}{\alpha}\to
\tyPrim{real}{\alpha\cdot\alpha}
\]
This captures the fact that the area of a circle varies as the square
of its radius, i.e., for any $s>0$, $\mathrm{areaCircle}\;(s\cdot r) =
s^2\cdot \mathrm{areaCircle}\;r$.  Here, $\alpha$ is interpreted as
the \emph{units of measure} of the argument to $\mathrm{areaCircle}$,
and `$\cdot$' composes units using the product. We can also add an
inverse operation and identity unit of measure, and then impose the
algebraic laws of abelian groups. This permits identification of, for
example, $\tyPrim{real}{\alpha\cdot\alpha^{-1}}$ with the type
$\tyPrim{real}{1}$ of dimensionless constants.

\paragraph{Abstraction}
In his original paper on parametricity, Reynolds asserted that
\emph{type structure is a syntactic discipline for enforcing levels of
  abstraction}.  We see something analogous here: if all primitive
operations are given types that reflect their behaviour under
translation, then there is no way to `break' this property. For
example, there is no way that $\mathrm{areaTri}$ can depend on the
actual coordinates of its inputs. Furthermore, the distinction between
points and vectors that is often enforced through abstract data
types~\cite{CGAL} is captured here by indices instead. For example,
the operation that takes two points and computes their vector
`difference' can be assigned the type
$\forall\alpha\mathord:\SynTransl.\tyPrim{vec}\alpha\times\tyPrim{vec}\alpha\to\tyPrim{vec}0$,
reflecting the invariance of the result (a pure vector) under
translations of the point arguments. As a result through types alone
we can, in essence, derive so-called \emph{coordinate-free}
geometry~\cite{CFGADT}.

The invariance properties discussed above can be seen as ``free
theorems''~\cite{wadler89theorems}, but the abstraction afforded by
polymorphic indexed types can also induce interesting type
\emph{isomorphisms}.  The type of $\mathrm{areaCircle}$ above is in
fact isomorphic to $\tyPrim{real}{1}$. A moment's thought reveals why:
what possible unary functions can be constructed whose outputs scale
as the square of the scaling of their inputs?  Answer: just those
functions of the form~$\lambda x. k x^2$ for some constant~$k$.  In
this case, of course, we expect that~$k = \pi$.

\paragraph{Relational parametricity}
To derive such invariance and abstraction properties of types, we
adopt the techniques of relational parametricity. Over an underlying
index-erasure semantics we construct binary relations parameterised by
an environment $\rho$ that describes how values of primitive type are
related according to their indices.  For example, values $v$ and $w$
of type $\tyPrim{real}\alpha$ are related when $v$ ``scales to'' $w$
according to an interpretation of $\alpha$ (i.e., $w=\rho(\alpha)\cdot
v$).  Values of polymorphic type are related exactly when they are
related for all possible interpretations of the quantified
variable. For example, values $v$ and $w$ of type
$\forall\alpha\mathord:\SynTransl.\tyPrim{vec}\alpha\to
\tyPrim{vec}\alpha$ are related when they are related at type
$\tyPrim{vec}\alpha\to\tyPrim{vec}\alpha$ for all translations
$t\in\Transl{2}$ associated with $\alpha$.

\vspace*{0.1in}

As it happens, the above examples can all be expressed using unary
functions rather than binary relations: we can write functions that
scale, or translate, one value to another according to their type,
even for higher-order function types. Moreover, the interpretation of
indices can be built up compositionally, by recursion over the
syntactic structure. For example, we have $\rho(\alpha\cdot\alpha) =
\rho(\alpha)\cdot\rho(\alpha)$ where product of units-of-measure is
interpreted by product of scale factors. For more sophisticated
applications, however, we need the fully generality of binary
relations. First, as is also the case with type polymorphism, it
provides the appropriate meaning for nested quantifiers in types such
as
$(\forall\alpha\mathord:\SynScal.\tyPrim{real}\alpha\to\tyPrim{real}\alpha)\to\tyBool$.
More interestingly, particular applications sometimes require
relations that are not built compositionally---we can, for instance,
use such relations to prove that there are no non-trivial functions of
type
$\forall\alpha\mathord:\SynScal.\tyPrim{real}{\alpha\cdot\alpha}\to\tyPrim{real}\alpha$.
Finally, there are interesting applications in which even the basic
relations are not functional. In a type system in which the index in
$\tyPrim{real}\alpha$ is interpreted not as a unit of measure, but as
a measure of \emph{closeness}, two values $x$ and $y$ of this type are
related if $|x-y| < \rho(\alpha)$ for a positive real number
$\rho(\alpha)$.  Rather beautifully, the standard notion of continuous
functions can then be expressed as a type:
\begin{displaymath}
  \forall \epsilon \mathord: \mathsf{R}^{>0}.\ \exists \delta\mathord: \mathsf{R}^{>0}.\ \tyPrim{real}{\delta} \to \tyPrim{real}{\epsilon}
\end{displaymath}

@@

This paper makes the following specific contributions:
\begin{itemize}
\item 
We present a collection of compelling examples of algebraically
indexed types, including a novel type system for geometry, a type
system for tracking taint marking (or provenance) of strings, a
refined type system for information flow based on logic, and a simple
type system for basic continuity analysis.
\item 
We formulate a type system that can either be used as a programming
language in its own right, or as the target of type-based
analyses. The type system consists of the usual type constructors
together with a collection of indexed primitive types, universal and
existential quantification over the indices, and a multi-sorted
equational theory for indices.
\item
We describe a relational semantics for the type system and prove an
analogue of Reynolds' Abstraction Theorem. We give precise conditions
on sets of relation environments under which our Abstraction Theorem
holds.
\item
For each of our main examples we deduce free theorems that are
consequences of our Abstraction Theorem, prove non-definability
results, and derive interesting type isomorphisms.
\end{itemize}
We have fully formalised our framework and Abstraction Theorem in Coq,
using strongly-typed term representations
throughout~\cite{TypedSyntax}.
